---
title: "The Four Causes of Adaptation"
subtitle: "How do students construct biological explanations?"
author: "Lucy E Delaney"
institute: "University of Illinois at Chicago"
date: "`r Sys.Date()`"
event: "SABER West, 2021"
output:
  xaringan::moon_reader:
    seal: false
    css:
      - css/manual-fonts.css
      - css/mytheme.css
    includes:
      after_body: css/insert-logo.html
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, echo=FALSE}
library(metathis)
library(htmltools)
```

class: right, middle, bg_cherry
# The Four Causes of `r htmltools::withTags(br())` Adaptation

`r htmltools::withTags(br())`

.pull-right[
.border[
We do not have knowledge of a thing until we have grasped its `r htmltools::withTags(b("why"))`, that is to say, its `r htmltools::withTags(b("cause"))`.]

Aristotle, `r htmltools::withTags(br())` _Physics: Book II_]

.pull-bottom[`r rmarkdown::metadata$author` · `r rmarkdown::metadata$event`]

???
hi there, i'm so glad to be here. today i'm going to talk to you about a project that i started in the Fall, examining how students think about complex questions regarding natural selection & adaptation.

---
class: left, middle, bg_munich

.pull-left[## Lucy Delaney 

`r htmltools::withTags(img(style = "border-radius: 50%;", src = "https://ledelaney.org/talks/sabertalk/css/images/Delaney-Headshot-Original.jpg", width = "170px"))`

[`r fontawesome::fa("envelope")` ldelan5@uic.edu](mailto:ldelan5@uic.edu)  
[`r fontawesome::fa("twitter")` @emergentlyLucy](https://twitter.com/emergentlyLucy)  
[`r fontawesome::fa("github")` @ledelaney](https:://github.com/ledelaney)]

.pull-right[## About me

`r emo::ji("mortar_board")` .medium-font[PhD candidate in Evolutionary Biology]
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r emo::ji("hibiscus")` .medium-font[Studying plant sex macroevolution]
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r emo::ji("hatching_chick")` .medium-font[Very first project in BER]]

.pull-bottom[
Link to all slide materials: https://github.com/ledelaney/01-21-SABERwest]

???
real quick: my name is lucy, i'm a phd candidate in evolutionary biology. i'm going to fly through this today: but, more materials online (these slides and some resources), so plz feel free to reach out!

---
class: left, bg_totoro

???
don't need to explain to this crowd how important understanding biological concepts is: can foster care&appreciation for the natural world, help to create empathetic&engaged citizens, bolster scientific&computational literacy... BUT...

--

## Biology education has a problem
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`

--

🥱
Biology is `r htmltools::HTML("&ldquo;")`just memorization!`r htmltools::HTML("&rdquo;")`

???
i hear this regularly from my students: often feels this way, because biology is usually presented as a string of facts, devoid of connection! each course siloed as if it's a discipline all its own

--

😫
Why does this matter: `r htmltools::HTML("&ldquo;")`I just want to be a doctor!`r htmltools::HTML("&rdquo;")`


???
i also here critiques such as this: this suggests we're not always doing a great job demonstrating the value of a robust intellectual framework in biology!

--
🤯
Formal instruction is not always leading to robust conceptual understanding

???
for me teaching genetics, surprising pauctiy of aha moments happening -- students are having a hard time putting the facts together and finding the bigger picture...
i think we need to be providing a better framework and a better narrative structure throughout the course of study

--

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
.center[
.large-font[
.pink[Evolutionary theory: help!]]]


???
the thing is, biology HAS a beautiful and compelling framework and narrative! and a lot of work over the last few decades suggests that evolution-centered teaching can help with these issues

---
class: middle, center, bg_fish

--

.border-opaque[
Despite its centrality to understanding biological complexity and its key practical relevance to medicine, biotechnology, and agriculture, natural selection remains one of the most widely misunderstood concepts of contemporary science.]
`r htmltools::withTags(br())`

.pull-right[
Keleman, 2012
`r htmltools::withTags(br())`
`r htmltools::withTags(em("Teleological minds"))`]


---
class: middle, left, bg_peacock


--
 `r emo::ji("yarn")` How do students construct biological explanations across levels of instruction?  


--
 `r emo::ji("chart_with_upwards_trend")` Does instruction improve outcomes?
 

???
first: four classes from 100- to 400-level
gave two weeks to answer five open-ended questions for homework, allowed to use resources and cite
completed once in the first two weeks of classes and once in the last
coded responses based on the following framework...

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

--

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (functional)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Material Cause:"), br(), br(), "All organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Effiecient Cause:"), br(), br(), "DNA mutates, organisms sexually reproduce, recombiantion", br(), br())))),
 tr(th("ULTIMATE (evolutionary)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Final cause:"), br(), br(), "All organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal cause:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Tinbergen, 1963; Nesse, 2013]]

???
borrowed a framework created by Tinbergen and organized by Nesse used to examine behavioral adaptations

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (functional)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Material Cause:"), br(), br(), "All organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Effiecient Cause:"), br(), br(), "DNA mutates, organisms sexually reproduce, recombiantion", br(), br())))),
 tr(th("ULTIMATE (evolutionary)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Final cause:"), br(), br(), "All organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal cause:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Tinbergen, 1963; Nesse, 2013]]

???
borrowed a framework created by Tinbergen and organized by Nesse used to examine behavioral adaptations

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (functional)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Material Cause:"), br(), br(), "All organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Effiecient Cause:"), br(), br(), "DNA mutates, organisms sexually reproduce, recombiantion", br(), br())))),
 tr(th("ULTIMATE (evolutionary)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Final cause:"), br(), br(), "All organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal cause:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Tinbergen, 1963; Nesse, 2013]]

???
borrowed a framework created by Tinbergen and organized by Nesse used to examine behavioral adaptations

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (functional)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Material Cause:"), br(), br(), "All organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Effiecient Cause:"), br(), br(), "DNA mutates, organisms sexually reproduce, recombiantion", br(), br())))),
 tr(th("ULTIMATE (evolutionary)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Final cause:"), br(), br(), "All organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal cause:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Tinbergen, 1963; Nesse, 2013]]

???
borrowed a framework created by Tinbergen and organized by Nesse used to examine behavioral adaptations

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (functional)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Material Cause:"), br(), br(), "All organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Effiecient Cause:"), br(), br(), "DNA mutates, organisms sexually reproduce, recombiantion", br(), br())))),
 tr(th("ULTIMATE (evolutionary)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Final cause:"), br(), br(), "All organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 1", br(), b("Formal cause:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Tinbergen, 1963; Nesse, 2013]]

???
borrowed a framework created by Tinbergen and organized by Nesse used to examine behavioral adaptations

---
class: middle, left, bg_ideal_v2

.pull-right[
We would love to see this after instruction!
]

???
this is what we HOPE to see after instruction....

---
class: middle, left, bg_reality_v2

???
but, this is what we actually see....

---
class: middle, left, bg_lian

.pull-left[
## Preliminary Takeaways

↪️
Most common explanations invoke Final and/or Efficient causality (i.e., ultimate pattern or proximate process) and shift regularly around these between pre-post
`r htmltools::withTags(br())`

↪️
Formal instruction is not nudging students toward Formal causality
`r htmltools::withTags(br())`

↪️
Students invoking Formal causality in both assignments are more likely to have more biology credits (_p-value_ = 0.06)]

--
.pull-low[
.center[
.mezzo-forte-font[
🧵 Weave evolutionary theory `r htmltools::HTML("&amp;")` principles into _every_ class!]]]

---
class: middle, left, bg_2birds

--

## Questions

`r htmltools::withTags(ol(li("Best methods for analyzing categorical data of this kind?"), li("Any general feedback!")))`

--
.bottom[
.large-font[
.pink[Thank you!]]]