---
title: "The Four Causes of Adaptation"
subtitle: "How do students construct biological explanations?"
author: "Lucy E Delaney"
institute: "University of Illinois at Chicago"
date: "`r Sys.Date()`"
event: "SABER West, 2021"
output:
  xaringan::moon_reader:
    seal: false
    css:
      - css/manual-fonts.css
      - css/mytheme.css
    includes:
      after_body: css/insert-logo.html
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, echo=FALSE}
library(metathis)
library(htmltools)
```

class: right, middle, bg_cherry
# The Four Causes of `r htmltools::withTags(br())` Adaptation

`r htmltools::withTags(br())`

.pull-right[
.border[
We do not have knowledge of a thing until we have grasped its `r htmltools::withTags(b("why"))`, that is to say, its `r htmltools::withTags(b("cause"))`.]

Aristotle, `r htmltools::withTags(br())` _Physics: Book II_]

.pull-bottom[`r rmarkdown::metadata$author` Â· `r rmarkdown::metadata$event`]

???
hi there, i'm so glad to be here. today i'm going to talk to you about a project that i started in the Fall for my dissertation, examining how students think about complex questions regarding natural selection & adaptation.

---
class: left, middle, bg_munich

.pull-left[## Lucy Delaney 

`r htmltools::withTags(img(style = "border-radius: 50%;", src = "https://ledelaney.org/talks/sabertalk/css/images/Delaney-Headshot-Original.jpg", width = "170px"))`

[`r fontawesome::fa("envelope")` ldelan5@uic.edu](mailto:ldelan5@uic.edu)  
[`r fontawesome::fa("twitter")` @emergentlyLucy](https://twitter.com/emergentlyLucy)  
[`r fontawesome::fa("github")` @ledelaney](https:://github.com/ledelaney)]

.pull-right[## About me

`r emo::ji("mortar_board")` .medium-font[PhD candidate in Evolutionary Biology]
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r emo::ji("hibiscus")` .medium-font[Studying plant sex macroevolution]
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r emo::ji("hatching_chick")` .medium-font[Very first project in BER]]

.pull-bottom[
Link to all slide materials: https://github.com/ledelaney/01-21-SABERwest]

???
real quick: my name is lucy, i'm a phd candidate in evolutionary biology. i'm going to fly through this today: but, more materials online and i'm dropping a link in the chat here, so plz feel free to reach out if you have questions!

---
class: left, bg_totoro

???
don't need to explain to this crowd how important a robust understanding of biological concepts is: can foster care&appreciation for the natural world, help to create empathetic&engaged citizens, bolster scientific&computational literacy... BUT...

--

## Biology education has a problem
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`

.footnote[
Momsen _et al._, 2010]

--

ðŸ¥±
Biology is `r htmltools::HTML("&ldquo;")`just memorization!`r htmltools::HTML("&rdquo;")`

.footnote[
Momsen _et al._, 2010]

???
one of the most common critiques from students i hear is this: 
it often feels this way, because biology is usually presented as a string of facts, devoid of connection! each course siloed as if it's a discipline all its own

--

ðŸ˜«
Why does this matter: `r htmltools::HTML("&ldquo;")`I just want to be a doctor!`r htmltools::HTML("&rdquo;")`

.footnote[
Momsen _et al._, 2010]


???
i also here critiques such as this: 
this suggests to me that we're not always doing a great job demonstrating the value of a robust intellectual framework in biology!

--
ðŸ¤¯
Formal instruction is not always leading to robust conceptual understanding

.footnote[
Momsen _et al._, 2010; Coley `r htmltools::withTags("&amp;")` Tanner, 2015]

???
for me teaching genetics, surprising paucity of aha moments happening -- students are having a hard time putting the facts together and finding the bigger picture...

i think we need to be providing a better framework and a better narrative structure throughout the entire course of postsecondary study

--

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
.center[
.large-font[
.pink[Evolutionary theory: help!]]]

.footnote[
Momsen _et al._, 2010; Coley `r htmltools::withTags("&amp;")` Tanner, 2015; Hanisch `r htmltools::withTags("&amp;")` Eirdosh, 2020]

???
the thing is, biology HAS a beautiful and compelling framework and narrative! and a lot of work over the last few decades suggests that evolution-centered teaching can help with issues of both conceptual understanding and value...

BUT if we want to get to the summit that is understanding evolution there's a long path to traverse... no time to discuss each step in the journey, but going to focus on the last two

---
class: top, bg_fish

## Biology is a historical science
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`

.footnote[
Mayr, 1961; Chapleau _et al._, 1988; Abrams `r htmltools::withTags("&amp;")` Southerland, 2001; Hanisch `r htmltools::withTags("&amp;")` Eirdosh, 2020]

???
biology is of course an experimental science, but it is also a historical science -- variation is generated by changing a part of parts of something that already existed through the process of mutation or recombination. these processes do create from scratch.

--

.pull-left-narrow[
.large-font[
ðŸ¥“ ðŸ·]]
.pull-right-wide[
`r htmltools::withTags(br())`
Need to distinguish pattern from generating process]

???
so given this, students first need to be able to distinguish a pattern from its generating process.


--

.pull-left-narrow[
.large-font[
ðŸª ðŸ”ª]]

.pull-right-wide[
`r htmltools::withTags(br())`
ï¸Need to evaluate both proximate and ultimate processes responsible for pattern]


???
BUT, not only can different processes produce the same pattern, but there are also both "proximate" (or functional) and "ultimate" (evolutionary) processes responsible for a given pattern. for instance, there's the individual that makes the scarf -- but why did they make the scarf? presumably because we have necks and those necks get cold.

and research shows that in individual biology courses we spend a tremendous amount of time on the "proximate" or "functional" processes and proportionally less time on the "ultimate" or "evolutionary" processes. this makes conceptual understanding of evolution -- and i would argue, biology in general! -- nearly impossible.

given these two final stages of conceptual understanding -- pattern vs process and proximate vs ultimate causality -- i set out to answer some questions

---
class: middle, left, bg_peacock


--
 `r emo::ji("hammer_and_wrench")` How do students construct biological explanations across levels of instruction?  


--
 `r emo::ji("chart_with_upwards_trend")` Does instruction improve outcomes?
 

???
How do students construct biological explanations across levels of instruction?  (used four classes from 100-400 level)

Does instruction improve outcomes? (pre-post design)

students had two weeks to answer five open-ended questions for homework, allowed to use resources -- once in the beginning of the semester and once at the end
coded responses based on the following framework...

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

--

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (functional)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Material Cause:"), br(), br(), "Organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Efficient Cause:"), br(), br(), "DNA mutates, many organisms sexually reproduce, recombination", br(), br())))),
 tr(th("ULTIMATE (evolutionary)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Final cause:"), br(), br(), "Organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal cause:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Aristotle; Tinbergen, 1963; Nesse, 2013]]

???
slightly tweaked a framework originally created by Niko Tinbergen that is most commonly used to examine behavioral adaptations

here is an example of one of the assignment questions: Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (functional)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Material Cause:"), br(), br(), "Organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Efficient Cause:"), br(), br(), "DNA mutates, many organisms sexually reproduce, recombination", br(), br())))),
 tr(th("ULTIMATE (evolutionary)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Final cause:"), br(), br(), "Organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal cause:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Aristotle; Tinbergen, 1963; Nesse, 2013]]

???
let's talk through the proximate/functional (or individual-level) row first:
-material cause addresses proximate-pattern: all organisms have different DNA, constructed differently

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (functional)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Material Cause:"), br(), br(), "Organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Efficient Cause:"), br(), br(), "DNA mutates, many organisms sexually reproduce, recombination", br(), br())))),
 tr(th("ULTIMATE (evolutionary)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Final cause:"), br(), br(), "Organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal cause:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Aristotle; Tinbergen, 1963; Nesse, 2013]]

???
-efficient cause is the proximate-process: DNA mutates, sexual reproduction/recombination

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (functional)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Material Cause:"), br(), br(), "Organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Efficient Cause:"), br(), br(), "DNA mutates, many organisms sexually reproduce, recombination", br(), br())))),
 tr(th("ULTIMATE (evolutionary)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Final cause:"), br(), br(), "Organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal cause:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Aristotle; Tinbergen, 1963; Nesse, 2013]]

???
now the ultimate/evolutionary (or species-level) row:
-final cause ultimate-pattern (dangerously teleological, but that's a discussion for another time): all organisms take advantage of the environment in different ways

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (functional)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Material Cause:"), br(), br(), "Organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Efficient Cause:"), br(), br(), "DNA mutates, many organisms sexually reproduce, recombination", br(), br())))),
 tr(th("ULTIMATE (evolutionary)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Final cause:"), br(), br(), "Organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 1", br(), b("Formal cause:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Aristotle; Tinbergen, 1963; Nesse, 2013]]

???
-formal cause ultimate-process (THE ANSWER WE WANT!): environmental heterogeneity + biological tradeoffs over large timescales

and what we were looking for in the students' responses is the FORMAL cause -- the ultimate process responsible, the full evolutionary history of the trait and organism

coded responses based on what level of explanation the student addressed in response (sometimes more than one)

---
class: middle, left, bg_ideal_v2

.pull-right[
We would love to see this after instruction!
]

???
so here is an example of some idealized data:
R1 response 1, R2 response 2... and the idealized case is that after formal instruction they are all nailing formal causality in their responses

---
class: middle, left, bg_reality_v2

???
but this is what we really see... this is currently binned by the number of earned biology credits... and it's a bit of a mess! there's no real clear pattern here: many students actually move away from formal causality after instruction or more often never get there at all

---
class: middle, left, bg_lian

.pull-left[
## Preliminary Takeaways

â†ªï¸
Most common explanations invoke Final and/or Efficient causality (i.e., ultimate-pattern or proximate-process) and shift regularly around these between pre-post
`r htmltools::withTags(br())`

â†ªï¸
Formal instruction is not nudging students toward Formal causality
`r htmltools::withTags(br())`

â†ªï¸
Students invoking Formal causality in both assignments are more likely to have more biology credits (_p-value_ = 0.06)]

--
.pull-low[
.center[
.mezzo-forte-font[
ðŸ§µ Weave evolutionary theory `r htmltools::HTML("&amp;")` principles into _every_ class!]]]

.footnote[
ArauÌjo, 2020]

???
MUST WEAVE EVOLUTIONARY THEORY INTO EACH CLASS:
1. studies show that narratives or stories are far easier to remember than disjointed facts
2. moreover, how can students form a conceptual framework if we neglect half of underlying causality?

---
class: middle, left, bg_2birds

--

## Questions

`r htmltools::withTags(ol(li("Best methods for analyzing categorical data of this kind?"), li("Any general feedback!")))`

--
.bottom[
.large-font[
.pink[Thank you!]]]

???
1. Best methods for analyzing categorical data of this kind? --> also have some educational history stuff, more of a quantitative person so this is all new to me
2. and the field is new to me as well, so Any general feedback! --> this is a dissertation chapter, and not necessarily something i plan on publishing in full, but would love any thoughts!